---
title: "Data Import"
output: github_document
---


```{r, echo = FALSE}
library(tidyverse)
library(readxl) 
## use library(readxl) to import package to import in excel files
library(haven)
## use library(haven) to import package to import SAS datasets
```

## Data Import:CSVs

Let's import data using the `readr` package

```{r}
litters_df = read_csv("data_import_examples/FAS_litters.csv")
#naming the file as litters dataframe

litters_df = janitor::clean_names(litters_df)
#janitor package 
#clean_names function within janitor package
#prevents from loading to many names
```

Look at the data
```{r}
litters_df

#first six rows
head(litters_df)

#last six rows
tail(litters_df)

view(litters_df)
#prints entire dataset
#use with eval=false so it doesn't break when knitting

skimr::skim(litters_df)
#quick summary of entire dataset
```

```{r, eval = FALSE}
#eval = FALSE

```

`read.csv` options..
col_names = TRUE - first row are the column headers

```{r}
read_csv("data_import_examples/FAS_litters.csv", na = c("", "NA", 999, 88), skip = 2)
#telling R it is missing
#skip = "#" - telling R to skip certain number of rows
```
<chr> = character variable
<dbl> = number variable

## col_types = --> use to correct variable type if R has incorrectly labeled them

##other file formats

We need to read in an excel spreadsheet...

```{r}
mlb_df = read_excel("data/mlb11.xlsx")
mlb_df = read_excel("data/mlb11.xlsx", range = "B3:B6")
#adding range will only print out those cells
mlb_df

lotr_words_df = read_excel("data/LotR_Words.xlsx")
```

## do not commit the files that are temporarily created

##importing SAS dataset

```{r}
pulse_df = read_sas("data/public_pulse_Data.sas7bdat")
pulse_df
```

##data export
```{r}
write_csv(lotr_words_df, file = "lotr_words_df.csv")
```



